Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's \emph{Irises}

### 2020-08-04 penguins.Rmd
update penguins paper

In 1936, one year after appearing in the Bulletin of the American Iris Society, Anderson’s *Iris* measurements were published by eugenicist and statistician Ronald A. Fisher in Annals of Eugenics [@fisher_use_1936]. Regardless of Anderson's initial motivation, the data today remain inextricably linked to Fisher’s eugenics research and are even commonly, if unfairly, referred to as "Fisher's iris data.” For example, an August 2020 search for “Iris flower data set” in Wikipedia returns an article beginning with “The iris flower data set or Fisher’s iris data…”, then later in the article: “It is sometimes called Anderson's Iris data set” [@wikipedia_iris_2020]. The *Iris* dataset is similarly credited to Fisher in statistical computing literature [@trendafilov_simple_2009, @wang_matlab_2015, @woods_how_2015, @chen_unsupervised_2018].

@@

      `remotes::install_github("allisonhorst/palmerpenguins")`.

@@

### Continuous quantitative variables: overview

Distributions, relationships between variables and clustering can be visually explored between species for the four structural size measurements in **iris** (sepal width and length, petal width and length) and **penguins** (flipper length, body mass, bill length and depth). It is worth noting that the correlation coefficients presented are Pearson’s *r*, developed by and named for another famous eugenicist and statistician, Karl Pearson [ref].

### Linear correlation, regression, and Simpson’s Paradox

There exist numerous opportunities to explore linear relationships and correlations in penguins and iris both within and across species (Figures 1 & 2). Here, we highlight one comparison that is uniquely similar: petal dimensions (petal length versus petal width) from iris, and penguin size (flipper length versus body mass) in penguins (Figure 3). The overall trend across all three species is approximately linear for both iris and penguins (Figure 3A,B). Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted (Figure 3A,B), compared to multiple linear regression with species included (Figure 3C,D).
Notably, distinctions between species are clearer for iris petals - particularly, the much smaller petals for Iris setosa - compared to penguins, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.

@@

Here, we have shown that the **penguins** data, available in the **palmerpenguins** R package, can replace **iris** for a number of common use cases in data science and statistics education including exploratory data visualization, linear correlation and regression, PCA and clustering by k-means. In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.

>>>>

In 1936, one year after being described in the Bulletin of the American Iris Society, Anderson’s *Iris* measurements were published in full by eugenicist and statistician Ronald A. Fisher in *Annals of Eugenics* [@fisher_use_1936]. Regardless of Anderson's initial motivation, the data today remain inextricably linked to Fisher’s eugenics research and are even commonly, if unfairly, referred to as "Fisher's iris data.” For example, an August 2020 search for “Iris flower data set” in Wikipedia returns an article beginning with “The iris flower data set or Fisher’s iris data…”, then later in the article: “It is sometimes called Anderson's Iris data set” [@wikipedia_iris_2020]. The *Iris* dataset is similarly credited to Fisher in statistical computing literature [@trendafilov_simple_2009, @wang_matlab_2015, @woods_how_2015, @chen_unsupervised_2018].

@@

      `remotes::install_github("allisonhorst/palmerpenguins")`

@@

```{r, echo = FALSE}
iris_counts <- iris %>% 
  count(Species) %>% 
  rename(`Iris species` = Species)

penguin_counts <- penguins %>% 
  mutate(species = as.character(species)) %>% 
  mutate(species = case_when(
    species == "Adelie" ~ "Adélie", 
    species == "Chinstrap" ~ "chinstrap",
    species == "Gentoo" ~ "gentoo",
    TRUE ~ species
  )) %>% 
  mutate(sex = str_to_title(sex)) %>% 
  group_by(species, sex) %>% 
  tally() %>% 
  pivot_wider(names_from = sex, values_from = n) %>% 
  replace_na(list(`NA` = 0)) %>% 
  as.data.frame() %>% 
  rename(`Penguin species` = species)

iris_penguin_n <- cbind(iris_counts, penguin_counts) %>%
   rename(`Sample size` = n)

iris_penguin_n %>% 
   kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.", align = "lclccc") %>% 
  add_header_above(c("Iris sample size (by species)" = 2, "Penguin sample size (by species and sex)" = 4 ))

```

### Continuous quantitative variables

Distributions, relationships between variables and clustering can be visually explored between species for the four structural size measurements in **penguins** (flipper length, body mass, bill length and depth; Figure \ref{fig:penguins-pairs}) and **iris** (sepal width and length, petal width and length; Figure \ref{fig:iris-pairs}).

```{r, penguins-pairs, fig.cap = "Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).", warning = FALSE, message = FALSE, echo = FALSE}
penguin_pairs <- penguins %>%
  mutate(species = as.character(species)) %>% 
  mutate(species = case_when(
    species == "Adelie" ~ "Adélie", 
    TRUE ~ species
  )) %>% 
  select(species, where(is.numeric)) %>% 
  ggpairs(aes(color = species, shape = species),
          columns = c("flipper_length_mm", "body_mass_g", 
                      "bill_length_mm", "bill_depth_mm"),
          columnLabels = c("Flipper length (mm)","Body mass (g)", "Bill length (mm)", "Bill depth (mm)"),
          upper = list(continuous = wrap("cor", size = 2.7, color = "black")),
          lower = list(continuous = wrap(ggally_points, size = 1.3))) +
  scale_fill_manual(values = c("#e59f00","#0072b2","#009e73")) +
  scale_color_manual(values = c("#e59f00","#0072b2","#009e73")) +
  scale_shape_manual(values = c(15,16,17)) +
  theme_minimal() +
  theme(
        text = element_text(size = 9),
        panel.grid.major = element_line(colour = NA),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "gray80", fill = NA)
        )

penguin_pairs
```

```{r, iris-pairs, fig.cap = "Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).", warning = FALSE, message = FALSE, echo = FALSE}
iris_pairs <- iris %>%
  ggpairs(aes(color = Species, shape = Species),
          columns = c("Petal.Length", "Petal.Width", 
                      "Sepal.Length", "Sepal.Width"),
          columnLabels = c("Petal length (cm)","Petal width (cm)", "Sepal length (cm)", "Sepal width (cm)"),
          upper = list(continuous = wrap("cor", size = 2.7, color = "black"))) +
  scale_colour_manual(values = c("gray70","gray40","black")) +
  scale_fill_manual(values = c("gray70","gray40","black")) +
  theme_minimal() +
  theme(
        text = element_text(size = 9),
        panel.grid.major = element_line(colour = NA),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "white", color = "gray80")
        )

iris_pairs

```

There exist numerous opportunities to explore linear relationships and correlations in **penguins** and **iris** both within and across species (Figures \ref{fig:penguins-pairs} & \ref{fig:iris-pairs}). Here, we highlight one comparison that is uniquely similar: petal dimensions (petal length versus petal width) from iris, and penguin size (flipper length versus body mass) in penguins (Figure 3). The overall trend across all three species is approximately linear for both iris and penguins (Figure 3A,B). Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted (Figure 3A,B), compared to multiple linear regression with species included (Figure 3C,D).

Notably, distinctions between species are clearer for iris petals - particularly, the much smaller petals for *Iris setosa* - compared to penguins, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.

@@

Educators have three options regarding **iris** use in course materials:

1. Use **iris** without addressing its use in Fisher's eugenics research
2. Use **iris** *and* address its use in Fisher's eugenics research
3. Replace **iris** with an alternative dataset offering similar teaching and learning opportunities

Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the **palmerpenguins** R package, can replace **iris** for a number of common use cases in data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means. In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.

The **iris** dataset is widespread today because it was employed by Fisher to advance statistical methods for eugenics research. While the *Irises* were described in Anderson's 1935 article in *The Bulletin of the American Iris Society* [@anderson_irises_1935], the measurements were published in full in Fisher's 1936 *Annals of Eugenics* publication [@fisher_use_1936] - a journal for which Fisher served as editor before becoming chair of the Department of Eugenics at University College London.

### 2020-08-04 penguins.tex
update penguins paper

In 1936, one year after appearing in the Bulletin of the American Iris
Society, Anderson's \emph{Iris} measurements were published by
eugenicist and statistician Ronald A. Fisher in Annals of Eugenics
\citep{fisher_use_1936}. Regardless of Anderson's initial motivation,
the data today remain inextricably linked to Fisher's eugenics research
and are even commonly, if unfairly, referred to as ``Fisher's iris
data.'' For example, an August 2020 search for ``Iris flower data set''
in Wikipedia returns an article beginning with ``The iris flower data
set or Fisher's iris data\ldots{}'', then later in the article: ``It is
sometimes called Anderson's Iris data set'' \citep{wikipedia_iris_2020}.
The \emph{Iris} dataset is similarly credited to Fisher in statistical
computing literature \citep[\citet{wang_matlab_2015},
\citet{woods_how_2015},
\citet{chen_unsupervised_2018}]{trendafilov_simple_2009}.

@@

\begin{verbatim}
  `remotes::install_github("allisonhorst/palmerpenguins")`. 
\end{verbatim}

@@

\hypertarget{continuous-quantitative-variables-overview}{%
\subsubsection{Continuous quantitative variables:
overview}\label{continuous-quantitative-variables-overview}}

Distributions, relationships between variables and clustering can be
visually explored between species for the four structural size
measurements in \textbf{iris} (sepal width and length, petal width and
length) and \textbf{penguins} (flipper length, body mass, bill length
and depth). It is worth noting that the correlation coefficients
presented are Pearson's \emph{r}, developed by and named for another
famous eugenicist and statistician, Karl Pearson {[}ref{]}.

\hypertarget{linear-correlation-regression-and-simpsons-paradox}{%
\subsubsection{Linear correlation, regression, and Simpson's
Paradox}\label{linear-correlation-regression-and-simpsons-paradox}}

There exist numerous opportunities to explore linear relationships and
correlations in penguins and iris both within and across species
(Figures 1 \& 2). Here, we highlight one comparison that is uniquely
similar: petal dimensions (petal length versus petal width) from iris,
and penguin size (flipper length versus body mass) in penguins (Figure
3). The overall trend across all three species is approximately linear
for both iris and penguins (Figure 3A,B). Teachers may encourage
students to explore how simple linear regression results and predictions
differ when the species variable is omitted (Figure 3A,B), compared to
multiple linear regression with species included (Figure 3C,D). Notably,
distinctions between species are clearer for iris petals - particularly,
the much smaller petals for Iris setosa - compared to penguins, in which
Adélie and chinstrap penguins are largely overlapping in body size (body
mass and flipper length), and are both generally smaller than gentoos.

@@

Here, we have shown that the \textbf{penguins} data, available in the
\textbf{palmerpenguins} R package, can replace \textbf{iris} for a
number of common use cases in data science and statistics education
including exploratory data visualization, linear correlation and
regression, PCA and clustering by k-means. In addition, we feel that
teaching and learning opportunities in \textbf{penguins} are increased
due to a greater number of variables, missing values, unequal sample
sizes, and Simpson's paradox examples.

>>>>

In 1936, one year after being described in the Bulletin of the American
Iris Society, Anderson's \emph{Iris} measurements were published in full
by eugenicist and statistician Ronald A. Fisher in \emph{Annals of
Eugenics} \citep{fisher_use_1936}. Regardless of Anderson's initial
motivation, the data today remain inextricably linked to Fisher's
eugenics research and are even commonly, if unfairly, referred to as
``Fisher's iris data.'' For example, an August 2020 search for ``Iris
flower data set'' in Wikipedia returns an article beginning with ``The
iris flower data set or Fisher's iris data\ldots{}'', then later in the
article: ``It is sometimes called Anderson's Iris data set''
\citep{wikipedia_iris_2020}. The \emph{Iris} dataset is similarly
credited to Fisher in statistical computing literature
\citep[\citet{wang_matlab_2015}, \citet{woods_how_2015},
\citet{chen_unsupervised_2018}]{trendafilov_simple_2009}.

@@

\begin{verbatim}
  `remotes::install_github("allisonhorst/palmerpenguins")`
\end{verbatim}

@@

\begin{Schunk}
\begin{table}

\caption{\label{tab:unnamed-chunk-1}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.}
\centering
\begin{tabular}[t]{l|c|l|c|c|c}
\hline
\multicolumn{2}{c|}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
\cline{1-2} \cline{3-6}
Iris species & Sample size & Penguin species & Female & Male & NA\\
\hline
setosa & 50 & Adélie & 73 & 73 & 6\\
\hline
versicolor & 50 & chinstrap & 34 & 34 & 0\\
\hline
virginica & 50 & gentoo & 58 & 61 & 5\\
\hline
\end{tabular}
\end{table}

\end{Schunk}

\hypertarget{continuous-quantitative-variables}{%
\subsubsection{Continuous quantitative
variables}\label{continuous-quantitative-variables}}

Distributions, relationships between variables and clustering can be
visually explored between species for the four structural size
measurements in \textbf{penguins} (flipper length, body mass, bill
length and depth; Figure \ref{fig:penguins-pairs}) and \textbf{iris}
(sepal width and length, petal width and length; Figure
\ref{fig:iris-pairs}).

\begin{Schunk}
\begin{figure}
\includegraphics{penguins_files/figure-latex/penguins-pairs-1} \caption[Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species]{Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:penguins-pairs}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}
\includegraphics{penguins_files/figure-latex/iris-pairs-1} \caption[Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species]{Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:iris-pairs}
\end{figure}
\end{Schunk}

There exist numerous opportunities to explore linear relationships and
correlations in \textbf{penguins} and \textbf{iris} both within and
across species (Figures \ref{fig:penguins-pairs} \&
\ref{fig:iris-pairs}). Here, we highlight one comparison that is
uniquely similar: petal dimensions (petal length versus petal width)
from iris, and penguin size (flipper length versus body mass) in
penguins (Figure 3). The overall trend across all three species is
approximately linear for both iris and penguins (Figure 3A,B). Teachers
may encourage students to explore how simple linear regression results
and predictions differ when the species variable is omitted (Figure
3A,B), compared to multiple linear regression with species included
(Figure 3C,D).

Notably, distinctions between species are clearer for iris petals -
particularly, the much smaller petals for \emph{Iris setosa} - compared
to penguins, in which Adélie and chinstrap penguins are largely
overlapping in body size (body mass and flipper length), and are both
generally smaller than gentoos.

@@

Educators have three options regarding \textbf{iris} use in course
materials:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Use \textbf{iris} without addressing its use in Fisher's eugenics
  research
\item
  Use \textbf{iris} \emph{and} address its use in Fisher's eugenics
  research
\item
  Replace \textbf{iris} with an alternative dataset offering similar
  teaching and learning opportunities
\end{enumerate}

Here, we have shown that structural size measurements for Palmer
Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins}
in the \textbf{palmerpenguins} R package, can replace \textbf{iris} for
a number of common use cases in data science and statistics education
including exploratory data visualization, linear correlation and
regression, PCA, and clustering by k-means. In addition, teaching and
learning opportunities in \textbf{penguins} are increased due to a
greater number of variables, missing values, unequal sample sizes, and
Simpson's paradox examples.

The \textbf{iris} dataset is widespread today because it was employed by
Fisher to advance statistical methods for eugenics research. While the
\emph{Irises} were described in Anderson's 1935 article in \emph{The
Bulletin of the American Iris Society} \citep{anderson_irises_1935}, the
measurements were published in full in Fisher's 1936 \emph{Annals of
Eugenics} publication \citep{fisher_use_1936} - a journal for which
Fisher served as editor before becoming chair of the Department of
Eugenics at University College London.
